\bf Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-06-25 paper.qmd
finished explanaiton of ellipse

---
title: "Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
---

@@

More generally, projection pursuit (REF) defines a quantitative criterion for the interestingness of a projection (a projection pursuit index), and searches the space of possible projections for the most interesting one to display. We can also define sequences of interpolated linear projections to better understand a multivariate distribution. Animating a randomly selected interpolated sequence of linear projections is called a grand tour (REF). The combination of these two approaches would then use a projection pursuit index to select interesting projections, but display them via an interpolated path to provide context. This is called a guided tour (REF).

@@

1. Simulating a sample of observations from $N_p(\mu, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean.

```{r}
#| echo: false
#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1. A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
set.seed(717)
d1 <- matrix(rnorm(716 * 2), ncol = 2) 
d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
evc <- eigen(vc)
vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
d3 <- d2 %*% vc2 |>
  as.data.frame()

>>>>

---
title: "Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
---

@@

More generally, projection pursuit (REF) defines a quantitative criterion for the interestingness of a projection (a projection pursuit index), and searches the space of possible projections for the most interesting one to display. We can also define sequences of interpolated linear projections to better understand a multivariate distribution. Animating a randomly selected interpolated sequence of linear projections is called a grand tour (REF). The combination of these two approaches would then use a projection pursuit index to select interesting projections, but display them via an interpolated path to provide context. This is called a guided tour (@cook1995).

@@

1. Simulating a sample of observations ($\mathbfit{x}$)
from $N_p(\mu, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.

To convert this to points on the surface of a confidence ellipse,

3. transform the shape using a specific variance covariance, and shift to center on the mean vector.

Finally, new observations can be visually compared with this ellipse by

4. plotting them together.

@fig-ci illustrates this process for \gD. This is easiest way to view this normal region relative to a new sample for any \pD{} problem.

```{r}
#| echo: false
#| label: fig-ci
#| fig-cap: "Simulating a uniform sample on a sphere involves sampling from a multivariate normal (a) and transforming each observation to have length equal to 1. A confidence ellipse is generated by transforming the sphere relative to a specified variance-covariance matrix (c), and new observations can be visually assessed to be inside or outside by plotting with the ellipse (d)."
#| out-width: 80%
set.seed(717)
d1 <- matrix(rnorm(716 * 2), ncol = 2) 
d2 <- t(apply(d1, 1, geozoo:::l2norm_vec))
vc <- matrix(c(1, -0.5, -0.5, 1), ncol=2, byrow=TRUE)
evc <- eigen(vc)
vc2 <- (evc$vectors) %*% diag(sqrt(evc$values)) %*% t(evc$vectors)
d3 <- d2 %*% vc2 |>
  as.data.frame()

### 2024-06-25 paper.tex
finished explanaiton of ellipse

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

More generally, projection pursuit (REF) defines a quantitative
criterion for the interestingness of a projection (a projection pursuit
index), and searches the space of possible projections for the most
interesting one to display. We can also define sequences of interpolated
linear projections to better understand a multivariate distribution.
Animating a randomly selected interpolated sequence of linear
projections is called a grand tour (REF). The combination of these two
approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (REF).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations from \(N_p(\mu, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean.
\end{enumerate}

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}

@@

\caption{Simulating a uniform sample on a sphere involves sampling from
a multivariate normal (a) and transforming each observation to have
length equal to 1. A confidence ellipse is generated by transforming the
sphere relative to a specified variance-covariance matrix (c), and new
observations can be visually assessed to be inside or outside by
plotting with the ellipse (d).}

@@

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

More generally, projection pursuit (REF) defines a quantitative
criterion for the interestingness of a projection (a projection pursuit
index), and searches the space of possible projections for the most
interesting one to display. We can also define sequences of interpolated
linear projections to better understand a multivariate distribution.
Animating a randomly selected interpolated sequence of linear
projections is called a grand tour (REF). The combination of these two
approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (\citet{cook1995}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations (\(\mathbfit{x}\)) from
  \(N_p(\mu, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean,
  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
\end{enumerate}

To convert this to points on the surface of a confidence ellipse,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  transform the shape using a specific variance covariance, and shift to
  center on the mean vector.
\end{enumerate}

Finally, new observations can be visually compared with this ellipse by

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  plotting them together.
\end{enumerate}

Figure~\ref{fig-ci} illustrates this process for \gD. This is easiest
way to view this normal region relative to a new sample for any \pD{}
problem.

\begin{figure}

\centering{

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-ci-1.pdf}

@@

\caption{\label{fig-ci}Simulating a uniform sample on a sphere involves
sampling from a multivariate normal (a) and transforming each
observation to have length equal to 1. A confidence ellipse is generated
by transforming the sphere relative to a specified variance-covariance
matrix (c), and new observations can be visually assessed to be inside
or outside by plotting with the ellipse (d).}

@@

\renewcommand\refname{Conclusion}
  \bibliography{bibliography.bib}
